---
title: "Tuition Data Creation and Cleaning"
author: "Matt Defenderfer"
date: "2/22/2022"
output: html_document
---

```{r libraries}
library(janitor)
library(tidyverse)
```


This document shows how the dataset used in the Intro to R Workshop during Research Computing Day 2022 was created. Multiple pieces of data individual data were cleaned and combined together. The pieces of data are as follows:

1. `tuition_cost` data provided by the March 3, 2020 entry of Tidy Tuesday. This entry can be found at https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-10/readme.m


##  Cost of Attendence

```{r read CoA data}
# read in data, clean variable names, remove unit id, institution alias, street address, and city, and finally rename institution and state variables
cost_attendance_raw <- read_csv('cost-attendance.csv') %>%
  clean_names() %>%
  select(-institution_name_alias_hd2016, -street_address_or_post_office_box_hd2016, -city_location_of_institution_hd2016) %>%
  rename(name = institution_name,
         state_code = state_abbreviation_hd2016)
```

```{r}
# Separate year from type of cost and store type of cost in its own column 'measure'
cost_attendance <- cost_attendance_raw %>%
  pivot_longer(cols = contains(c('total','published')), 
               names_to = c('measure','.value'), 
               names_pattern = '(.*)_(20[\\d]{2}_[\\d]{2}).*', 
               values_to = 'cost')
```

```{r}
# separate the measure column into three separate variables of cost type (tuition+fees only vs. total), in-state or out-of-state, and living with parents or not
cost_attendance <- cost_attendance %>%
  mutate(cost_type = if_else(str_detect(measure,'published'),'tuition','total'),
         
         in_state = if_else(str_detect(measure,'in_state'), 'in', 'out'),
         
         on_campus = case_when(
           str_detect(measure,'on_campus') ~ 'on',
           str_detect(measure,'off_campus') ~ 'off'),
         
         living_with_family = case_when(
           str_detect(measure,'not_with_family') ~ 'without',
           str_detect(measure,'with_family') ~ 'with')) %>%
  relocate(cost_type:living_with_family, .after = measure) %>%
  select(-measure)

# remove off campus category to simplify the data. This also functionally removes the living_with_parents category since living with and without parents was predicated on the individual living off campus. Now, it will be in-state vs. out-of-state for on-campus students only
cost_attendance <- cost_attendance %>% 
  filter(on_campus %in% c('on',NA)) %>%
  select(-living_with_family,-on_campus)
```

```{r}
# rename year columns to be academic year. So 2017_18 will become AY2017
cost_attendance <- cost_attendance %>%
  rename_with(~str_remove(.,'_[\\d]{2}'), contains('20')) %>%
  rename_with(~paste0('AY',.), contains('20'))
```

```{r}
# remove AY2008, has missing data from at least one measure in every school.
cost_attendance <- cost_attendance %>%
  select(-AY2008)
```

```{r}
# only keep data that has values for all academic years for all categories
keep_schools <- cost_attendance %>%
  group_by(name) %>%
  summarize(across(contains('AY'),~all(!is.na(.)))) %>%
  filter(if_all(contains('AY'), ~ . == TRUE)) %>%
  select(name)

cost_attendance <- cost_attendance %>%
  filter(name %in% keep_schools$name)
```

## Degree Type and Economic Model Information

Some other interesting information is whether a university is a 2-year vs 4-year institution and whether it is public, private, or for-profit. These categories can be informative to general trends in tuition within each of these groups. Tuition Tracker originally downloads data from [IPEDS](https://nces.ed.gov/ipeds/use-the-data) which is where I downloaded these data from as well.

### Economic Model Information

```{r load net price data}
inst_info <- read_csv('institution-information.csv') %>%
  clean_names()

# rename some of the columns
inst_info <- inst_info %>%
  rename(name = institution_name,
         state_code = state_abbreviation_hd2016) %>%
  rename_with(~str_remove(.,'_of_institution_hd2016'), .cols = contains('_of_institution_hd2016'))
```

```{r}
# select the columns we want and change them to factors with their descriptions
inst_info <- inst_info %>%
  select(unit_id,name,state_code,level,control) %>%
  filter(level != 3) %>%
  mutate(level = factor(level, levels = c(1,2), labels = c('4 Year','2 Year')),
         control = factor(control, levels = c(1,2,3), labels = c('Public','Private','For Profit')))
```


```{r merge cost_attendance with inst_info}
cost_attendance <- inst_info %>%
  right_join(cost_attendance, by = c('unit_id','name','state_code'))
```


## Graduation Rate and Degree Type

```{r load graduation rate data}
graduation <- read_csv('grad-rates.csv') %>%
  clean_names() %>%
  select(unit_id,institution_name,state_abbreviation_hd2016,contains('grand_total')) %>%
  select(unit_id,institution_name,state_abbreviation_hd2016,contains('rate')) %>%
  rename(name = institution_name,
         state_code = state_abbreviation_hd2016)
```

```{r}
# remaining grad rate variables have 4-year or 2-year college information in them, but each college is only one type, so all colleges have half data as NA. Easiest way to remove these rows is to pivot the table and put the year+college type in a single column and the rate in another column. Can also easily extract year information only

graduation <- graduation %>%
  pivot_longer(contains('rate'), names_to = 'year', values_to = 'rate') %>%
  mutate(year = paste0('AY',str_extract(year, '20[\\d]{2}'))) %>%
  filter(!is.na(rate))
```


```{r}
# only want colleges with graduation rates for all 6 years, so pivot wider (also sets up a pivot example during the course if the data is wide to begin with) and filter for colleges that have data for all 6 years

graduation <- graduation %>%
  pivot_wider(names_from = year, values_from = rate) %>%
  filter(if_all(contains('AY'),~!is.na(.)))
```

## Write Cleaned Data

```{r write cleaned data}
# filter out some of the variables to make the data more succinct. Only need costs for on-campus students. Also remove all colleges who do not have data for all years
cost_attendance %>%
  filter(on_campus == 'on') %>%
  filter(if_all(where(is.numeric),~!is.na(.))) %>%
  select(-on_campus,-living_with_family)

# somehow removed only tuition cost from the table, need to go back and check where.

  
write_csv('../cleaned_data/cost_of_attendance.csv')
write_csv(graduation,'../cleaned_data/graduation_rate.csv')
```

